Estimating Carrying Capacity for Juvenile Salmon using Quantile Random Forest Models

### 2020-12-01 analysis/paper/paper.Rmd
for final submission, kept the .tex file, changed format of saved figures. Also added linesep = "" to all tables, and added list of figures

---
title: "Estimating Carrying Capacity for Juvenile Salmon using Quantile Random Forest Models"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [biomark]
      correspondence: true
  - Michael W. Ackerman:
      email: Mike.Ackerman@merck.com
      institute: [biomark]
      correspondence: false
  - Richard A. Carmichael:
      email: Richard.Carmichael@merck.com
      institute: [biomark]
      correspondence: false
  - Sarah L. Hoffmann:
      email: Sarah.Hoffmann@merck.com
      institute: [biomark]
      correspondence: false
  - Chris Beasley:
      email: Chris.Beasley@merck.com
      institute: [biomark]
      correspondence: false
institute:
  - biomark: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: false
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
      includes:
        in_header: ../templates/header_manuscript.tex
    bookdown::word_document2:
      fig_caption: yes
      # reference_docx: "../templates/EcologicalApplicationsTemplate.docx" # Insert path for the DOCX file
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: false
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::html_document2:
      fig_caption: yes
      number_sections: false
      toc: no
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
csl: "../templates/ecological-applications.csl" # Insert path for the bib-style
abstract: |
  Establishing robust methods and metrics to evaluate habitat quality is critical for the recovery of endangered Pacific salmonids. A variety of modeling approaches are used for status and trend monitoring of anadromous species throughout the Pacific Northwest, USA, but current methods may fail to capture the complex relationship between fish and habitat and are often limited in predictive power beyond specific watersheds. Further, the focus on species distribution and abundance is not easily manipulated to predict carrying capacity and traditional stock-recruitment analyses are reliant on long-term data which are not always available. In this study, we developed a quantile random forest (QRF) model to  provide estimates of habitat carrying capacity for Chinook salmon parr during the summer months, at both the site and watershed scale. QRF models allow for the consideration of noisy data, correlated variables, and non-linear relationships: common features in fish-habitat datasets. We leveraged Columbia Habitat Monitoring Program (CHaMP) data to select habitat co-variates and predict capacity at those sites. We also identified a set of globally available attributes to extrapolate capacity estimate predictions throughout wadeable streams within the Columbia River basin. Total capacity estimates for watersheds closely matched estimates from alternative fish productivity models. Carrying capacity estimates based on QRF, like those presented here, provide managers a framework to guide the identification, prioritization, and development of habitat rehabilitation actions to recover salmon populations.
keywords: |
  carrying capacity; quantile regression; random forests; Chinook; quantile random forest
highlights: |
  These are the highlights. 
---

@@

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 600
)
```

@@

theme_set(theme_pubr())

@@

```{r fish-hab-sites}
qrf_data %>%
  tabyl(Watershed) %>%
  adorn_totals() %>%
  adorn_pct_formatting() %>%
  rename(`n Sites` = n,
         Percent = percent) %>%
  kable(caption = 'The number of unique sites, by watershed, with paired fish-habitat data used to populate the spring/summer-run Chinook salmon parr capacity QRF model',
        booktabs = T) %>%
  kable_styling(latex_options = c("hold_position"))

@@

```{r select-mets-table, eval = T}
qrf_mod$importance %>%
  as_tibble(rownames = 'ShortName') %>%
  rename(imp = IncNodePurity) %>%
  arrange(desc(imp)) %>%
  mutate(Rank = 1:n()) %>%
  select(-imp) %>%
  left_join(hab_dict %>%
              select(-Name) %>%
              inner_join(sel_hab_mets %>%
                           rename(ShortName = Metric)) %>%
              mutate(Name = str_replace(Name, '\n', ' '))) %>%
  filter(MetricGroupName != 'Tier 1 Summary') %>%
  select(Rank,
         Metric = Name,
         `Metric Category` = MetricCategory,
         Description = DescriptiveText) %>%
  kable(caption = "Habitat metrics and descriptions of metrics included in the QRF model to predict spring/summer-run Chinook salmon parr capacity. Metrics are ranked in order of relative importance.",
        booktabs = T,
        format.args = list(width = c(2, 10, 10, 50))) %>%
  kable_styling(latex_options = c("hold_position")) %>%
  column_spec(4, width = '3in')

@@

```{r gaa-covars, eval = T}
tibble(Covariate = gaa_covars) %>%
  left_join(gaa_meta %>%
              select(Covariate = FieldName,
                     Name = DisplayName,
                     Units,
                     SpatialScale)) %>%
  mutate(Units = if_else(Units == "na",
                         as.character(NA),
                         Units),
         Units = if_else(grepl('\\^2', Units),
                         str_replace(Units, '\\^2', '$^2$'),
                         Units)) %>%
  mutate(Name = if_else(Covariate == 'S2_02_11' & is.na(Name),
                        'NorWeST Aug. Temperature',
                        Name),
         Units = if_else(Covariate == 'S2_02_11',
                                'C',
                                Units),
         SpatialScale = if_else(Covariate == 'S2_02_11',
                                'Reach-2km',
                                SpatialScale)) %>%
  left_join(model_svy_df$mod_full[[1]] %>%
              summary() %>%
              coefficients() %>%
              as_tibble(rownames = "Covariate") %>%
              select(1:3)) %>%
  select(-Covariate) %>%
  rename(`Spatial Scale` = SpatialScale,
         Covariate = Name) %>%
  kable(caption = "Globally available attribute (GAA) habitat covariates used to extrapolate quantile random forest (QRF) model predictions of spring/summer-run Chinook parr capacity to a larger scale (e.g., watershed, population), with their coefficients and standard errors.",
        booktabs = T, 
        escape = F,
        digits = 3) %>%
  kable_styling(latex_options = c("hold_position"))

@@

```{r extrap-r2-table, eval = T}
model_svy_df %>%
  ungroup() %>%
  mutate(n_full = map_dbl(data,
                          .f = nrow),
         k_full = map_dbl(mod_full,
                          .f = function(x) {
                            names(x$coefficients)[!grepl('Intercept', names(x$coefficients))] %>%
                              length
                          }),
         r_full = map_dbl(mod_full,
                          .f = function(x) {
                            cor(x$y, x$fitted.values)
                          }),
         r2_full = r_full^2,
         r2_adj_full = 1 - ( ( (1 - r2_full) * (n_full - 1) ) / (n_full - k_full - 1) )) %>%
  mutate(n_no_champ = map_dbl(data,
                              .f = nrow),
         k_no_champ = map_dbl(mod_no_champ,
                              .f = function(x) {
                                names(x$coefficients)[!grepl('Intercept', names(x$coefficients))] %>%
                                  length
                              }),
         r_no_champ = map_dbl(mod_no_champ,
                              .f = function(x) {
                                cor(x$y, x$fitted.values)
                              }),
         r2_no_champ = r_no_champ^2,
         r2_adj_no_champ = 1 - ( ( (1 - r2_no_champ) * (n_no_champ - 1) ) / (n_no_champ - k_no_champ - 1) )) %>%
  select(response, starts_with("r2")) %>%
  gather(type, r2, -response) %>%
  mutate(model = if_else(grepl('no_champ', type),
                         'non-CHaMP',
                         'CHaMP'),
         type = if_else(grepl('adj', type),
                        'Adjusted R2',
                        'R2')) %>%
  spread(type, r2) %>%
  mutate(response = recode(response,
                           'chnk_per_m' = 'fish/m',
                           'chnk_per_m2' = 'fish/m$^2$')) %>%
  filter(response == "fish/m") %>%
  select(Model = model,
         Response = response,
         `$r^2$` = R2, 
         `Adjusted $r^2$` = `Adjusted R2`) %>%
  kable(caption = "Summary of extrapolation model fits, split by whether the extrapolation model used CHaMP watershed as a covariate or not.",
        escape = F,
        booktabs = T,
        digits = 3) %>%
  kable_styling(latex_options = c("hold_position"))

@@

```{r sr-table, eval = T}
spawn_rec_params %>%
  select(Population, data) %>%
  unnest(data) %>%
  group_by(Population) %>%
  summarise(`n Yrs` = n_distinct(BroodYear),
            Spawner_type = unique(Spawner_type)) %>%
  select(Population, `n Yrs`, Spawner_type) %>%
  mutate(Spawner_type = recode(Spawner_type,
                               'total spawners' = 'Spawners',
                               'redd counts' = 'Redds')) %>%
  rename(`Adult Data` = Spawner_type) %>%
  mutate(`Parr Data` = if_else(grepl('Chiwawa', Population),
                               'Parr Surveys',
                               'RST')) %>%
  full_join(spawn_rec_params %>%
              mutate(prnt_val = paste0(prettyNum(round(est), big.mark = ','), ' (', round(cv, 3), ')')) %>%
              select(Population, form, prnt_val) %>%
              mutate(form = recode(form,
                                   'BevertonHolt' = 'Beverton Holt',
                                   'Hockey' = 'Hockey Stick')) %>%
              mutate(form = factor(form,
                                   levels = c('Beverton Holt',
                                              'Ricker',
                                              'Hockey Stick',
                                              'QRF'))) %>%
              spread(form, prnt_val)) %>%
  filter(Population %in% keep_wtsds) %>%
  mutate(Population = recode(Population,
                             'Chiwawa R.' = 'Chiwawa River')) %>%
  kable(caption = "Estimates of parr capacity from both spawner-recruit data (Beverton-Holt, Ricker, hockey stick) and from extrapolated estimates of parr capacity from the quantile random forest (QRF) model. Numbers in parentheses are coefficients of variation.",
        booktabs = T) %>%
  kable_styling(latex_options = c("hold_position")) %>%
  # kable_styling(latex_options = c('scale_down')) %>%
  column_spec(1, width = "1in") %>%
  column_spec(c(5:8), width = "0.5in")

@@

>>>>

---
title: "Estimating Carrying Capacity for Juvenile Salmon using Quantile Random Forest Models"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [biomark]
      correspondence: true
  - Michael W. Ackerman:
      email: Mike.Ackerman@merck.com
      institute: [biomark]
      correspondence: false
  - Richard A. Carmichael:
      email: Richard.Carmichael@merck.com
      institute: [biomark]
      correspondence: false
  - Sarah L. Hoffmann:
      email: Sarah.Hoffmann@merck.com
      institute: [biomark]
      correspondence: false
  - Chris Beasley:
      email: Chris.Beasley@merck.com
      institute: [biomark]
      correspondence: false
institute:
  - biomark: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: false
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      keep_tex: true
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
      includes:
        in_header: ../templates/header_manuscript.tex
    bookdown::word_document2:
      fig_caption: yes
      # reference_docx: "../templates/EcologicalApplicationsTemplate.docx" # Insert path for the DOCX file
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: false
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::html_document2:
      fig_caption: yes
      number_sections: false
      toc: no
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
csl: "../templates/ecological-applications.csl" # Insert path for the bib-style
abstract: |
  Establishing robust methods and metrics to evaluate habitat quality is critical for the recovery of endangered Pacific salmonids. A variety of modeling approaches are used for status and trend monitoring of anadromous species throughout the Pacific Northwest, USA, but current methods may fail to capture the complex relationship between fish and habitat and are often limited in predictive power beyond specific watersheds. Further, the focus on species distribution and abundance is not easily manipulated to predict carrying capacity and traditional stock-recruitment analyses are reliant on long-term data which are not always available. In this study, we developed a quantile random forest (QRF) model to  provide estimates of habitat carrying capacity for Chinook salmon parr during the summer months, at both the site and watershed scale. QRF models allow for the consideration of noisy data, correlated variables, and non-linear relationships: common features in fish-habitat datasets. We leveraged Columbia Habitat Monitoring Program (CHaMP) data to select habitat co-variates and predict capacity at those sites. We also identified a set of globally available attributes to extrapolate capacity estimate predictions throughout wadeable streams within the Columbia River basin. Total capacity estimates for watersheds closely matched estimates from alternative fish productivity models. Carrying capacity estimates based on QRF, like those presented here, provide managers a framework to guide the identification, prioritization, and development of habitat rehabilitation actions to recover salmon populations.
keywords: |
  carrying capacity; quantile regression; random forests; Chinook; quantile random forest
highlights: |
  These are the highlights. 
---

@@

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 400,
  dev = c("pdf", 'tiff')
)
```

@@

theme_set(theme_pubr(base_family = "Helvetica"))

@@

```{r fish-hab-sites}
qrf_data %>%
  tabyl(Watershed) %>%
  adorn_totals() %>%
  adorn_pct_formatting() %>%
  rename(`n Sites` = n,
         Percent = percent) %>%
  kable(caption = 'The number of unique sites, by watershed, with paired fish-habitat data used to populate the spring/summer-run Chinook salmon parr capacity QRF model',
        booktabs = T,
        linesep = "") %>%
  kable_styling(latex_options = c("hold_position"))

@@

```{r select-mets-table, eval = T}
qrf_mod$importance %>%
  as_tibble(rownames = 'ShortName') %>%
  rename(imp = IncNodePurity) %>%
  arrange(desc(imp)) %>%
  mutate(Rank = 1:n()) %>%
  select(-imp) %>%
  left_join(hab_dict %>%
              select(-Name) %>%
              inner_join(sel_hab_mets %>%
                           rename(ShortName = Metric)) %>%
              mutate(Name = str_replace(Name, '\n', ' '))) %>%
  filter(MetricGroupName != 'Tier 1 Summary') %>%
  select(Rank,
         Metric = Name,
         `Metric Category` = MetricCategory,
         Description = DescriptiveText) %>%
  kable(caption = "Habitat metrics and descriptions of metrics included in the QRF model to predict spring/summer-run Chinook salmon parr capacity. Metrics are ranked in order of relative importance.",
        booktabs = T,
        linesep = "",
        format.args = list(width = c(2, 10, 10, 50))) %>%
  kable_styling(latex_options = c("hold_position")) %>%
  column_spec(4, width = '3in')

@@

```{r gaa-covars, eval = T}
tibble(Covariate = gaa_covars) %>%
  left_join(gaa_meta %>%
              select(Covariate = FieldName,
                     Name = DisplayName,
                     Units,
                     SpatialScale)) %>%
  mutate(Units = if_else(Units == "na",
                         as.character(NA),
                         Units),
         Units = if_else(grepl('\\^2', Units),
                         str_replace(Units, '\\^2', '$^2$'),
                         Units)) %>%
  mutate(Name = if_else(Covariate == 'S2_02_11' & is.na(Name),
                        'NorWeST Aug. Temperature',
                        Name),
         Units = if_else(Covariate == 'S2_02_11',
                                'C',
                                Units),
         SpatialScale = if_else(Covariate == 'S2_02_11',
                                'Reach-2km',
                                SpatialScale)) %>%
  left_join(model_svy_df$mod_full[[1]] %>%
              summary() %>%
              coefficients() %>%
              as_tibble(rownames = "Covariate") %>%
              select(1:3)) %>%
  select(-Covariate) %>%
  rename(`Spatial Scale` = SpatialScale,
         Covariate = Name) %>%
  kable(caption = "Globally available attribute (GAA) habitat covariates used to extrapolate quantile random forest (QRF) model predictions of spring/summer-run Chinook parr capacity to a larger scale (e.g., watershed, population), with their coefficients and standard errors.",
        booktabs = T,
        linesep = "", 
        escape = F,
        digits = 3) %>%
  kable_styling(latex_options = c("hold_position"))

@@

```{r extrap-r2-table, eval = T}
model_svy_df %>%
  ungroup() %>%
  mutate(n_full = map_dbl(data,
                          .f = nrow),
         k_full = map_dbl(mod_full,
                          .f = function(x) {
                            names(x$coefficients)[!grepl('Intercept', names(x$coefficients))] %>%
                              length
                          }),
         r_full = map_dbl(mod_full,
                          .f = function(x) {
                            cor(x$y, x$fitted.values)
                          }),
         r2_full = r_full^2,
         r2_adj_full = 1 - ( ( (1 - r2_full) * (n_full - 1) ) / (n_full - k_full - 1) )) %>%
  mutate(n_no_champ = map_dbl(data,
                              .f = nrow),
         k_no_champ = map_dbl(mod_no_champ,
                              .f = function(x) {
                                names(x$coefficients)[!grepl('Intercept', names(x$coefficients))] %>%
                                  length
                              }),
         r_no_champ = map_dbl(mod_no_champ,
                              .f = function(x) {
                                cor(x$y, x$fitted.values)
                              }),
         r2_no_champ = r_no_champ^2,
         r2_adj_no_champ = 1 - ( ( (1 - r2_no_champ) * (n_no_champ - 1) ) / (n_no_champ - k_no_champ - 1) )) %>%
  select(response, starts_with("r2")) %>%
  gather(type, r2, -response) %>%
  mutate(model = if_else(grepl('no_champ', type),
                         'non-CHaMP',
                         'CHaMP'),
         type = if_else(grepl('adj', type),
                        'Adjusted R2',
                        'R2')) %>%
  spread(type, r2) %>%
  mutate(response = recode(response,
                           'chnk_per_m' = 'fish/m',
                           'chnk_per_m2' = 'fish/m$^2$')) %>%
  filter(response == "fish/m") %>%
  select(Model = model,
         Response = response,
         `$r^2$` = R2, 
         `Adjusted $r^2$` = `Adjusted R2`) %>%
  kable(caption = "Summary of extrapolation model fits, split by whether the extrapolation model used CHaMP watershed as a covariate or not.",
        escape = F,
        booktabs = T,
        linesep = "",
        digits = 3) %>%
  kable_styling(latex_options = c("hold_position"))

@@

```{r sr-table, eval = T}
spawn_rec_params %>%
  select(Population, data) %>%
  unnest(data) %>%
  group_by(Population) %>%
  summarise(`n Yrs` = n_distinct(BroodYear),
            Spawner_type = unique(Spawner_type)) %>%
  select(Population, `n Yrs`, Spawner_type) %>%
  mutate(Spawner_type = recode(Spawner_type,
                               'total spawners' = 'Spawners',
                               'redd counts' = 'Redds')) %>%
  rename(`Adult Data` = Spawner_type) %>%
  mutate(`Parr Data` = if_else(grepl('Chiwawa', Population),
                               'Parr Surveys',
                               'RST')) %>%
  full_join(spawn_rec_params %>%
              mutate(prnt_val = paste0(prettyNum(round(est), big.mark = ','), ' (', round(cv, 3), ')')) %>%
              select(Population, form, prnt_val) %>%
              mutate(form = recode(form,
                                   'BevertonHolt' = 'Beverton Holt',
                                   'Hockey' = 'Hockey Stick')) %>%
              mutate(form = factor(form,
                                   levels = c('Beverton Holt',
                                              'Ricker',
                                              'Hockey Stick',
                                              'QRF'))) %>%
              spread(form, prnt_val)) %>%
  filter(Population %in% keep_wtsds) %>%
  mutate(Population = recode(Population,
                             'Chiwawa R.' = 'Chiwawa River')) %>%
  kable(caption = "Estimates of parr capacity from both spawner-recruit data (Beverton-Holt, Ricker, hockey stick) and from extrapolated estimates of parr capacity from the quantile random forest (QRF) model. Numbers in parentheses are coefficients of variation.",
        booktabs = T,
        linesep = "") %>%
  kable_styling(latex_options = c("hold_position")) %>%
  # kable_styling(latex_options = c('scale_down')) %>%
  column_spec(1, width = "1in") %>%
  column_spec(c(5:8), width = "0.5in")

@@

\listoffigures
